\begin{Schunk}
\begin{Sinput}
 kemz.3 <- MARSS(dat.z, model = dfa.model, control = cntl.list)
\end{Sinput}
\begin{Soutput}
Warning! Reached maxit before parameters converged. Maxit was 50.
 neither abstol nor log-log convergence tests were passed.

MARSS fit is
Estimation method: kem 
Convergence test: conv.test.slope.tol = 0.5, abstol = 0.001
WARNING: maxit reached at  50  iter before convergence.
 Neither abstol nor log-log convergence test were passed.
 The likelihood and params are not at the MLE values.
 Try setting control$maxit higher.
Log-likelihood: -782.202 
AIC: 1598.404   AICc: 1599.463   
 
                            Estimate
Z.z11                         0.4163
Z.z21                         0.5364
Z.z31                         0.2780
Z.z41                         0.5179
Z.z51                         0.1611
Z.z22                         0.6757
Z.z32                        -0.2381
Z.z42                        -0.2381
Z.z52                        -0.2230
Z.z33                         0.2305
Z.z43                        -0.1225
Z.z53                         0.3887
R.(Cryptomonas,Cryptomonas)   0.6705
R.(Diatoms,Diatoms)           0.0882
R.(Greens,Greens)             0.7201
R.(Unicells,Unicells)         0.1865
R.(Other.algae,Other.algae)   0.5441
Initial states (x0) defined at t=0

Standard errors have not been calculated. 
Use MARSSparamCIs to compute CIs and bias estimates.

Convergence warnings
10 warnings. First 10 shown.  Type cat(object$errors) to see the full list.
 Warning: the  Z.z51  parameter value has not converged.
 Warning: the  Z.z32  parameter value has not converged.
 Warning: the  Z.z52  parameter value has not converged.
 Warning: the  Z.z33  parameter value has not converged.
 Warning: the  Z.z43  parameter value has not converged.
 Warning: the  R.(Diatoms,Diatoms)  parameter value has not converged.
 Warning: the  R.(Greens,Greens)  parameter value has not converged.
 Warning: the  R.(Other.algae,Other.algae)  parameter value has not converged.
 Warning: the  logLik  parameter value has not converged.
 Type MARSSinfo("convergence") for more info on this warning.
\end{Soutput}
\end{Schunk}
