\begin{Schunk}
\begin{Sinput}
 royale.model.1 <- list(
   Z = "identity", B = "unconstrained",
   Q = "diagonal and unequal", R = "diagonal and unequal",
   U = "zero", tinitx = 1
 )
 cntl.list <- list(allow.degen = FALSE, maxit = 200)
 kem.1 <- MARSS(z.royale.dat, model = royale.model.1, control = cntl.list)
\end{Sinput}
\begin{Soutput}
Warning! Reached maxit before parameters converged. Maxit was 200.
 neither abstol nor log-log convergence tests were passed.

MARSS fit is
Estimation method: kem 
Convergence test: conv.test.slope.tol = 0.5, abstol = 0.001
WARNING: maxit reached at  200  iter before convergence.
 Neither abstol nor log-log convergence test were passed.
 The likelihood and params are not at the MLE values.
 Try setting control$maxit higher.
Log-likelihood: -75.56383 
AIC: 171.1277   AICc: 173.4933   
 
                     Estimate
R.(Wolf,Wolf)        0.001433
R.(Moose,Moose)      0.000362
B.(1,1)              0.768031
B.(2,1)             -0.178990
B.(1,2)              0.078335
B.(2,2)              0.827922
Q.(X.Wolf,X.Wolf)    0.455214
Q.(X.Moose,X.Moose)  0.178661
x0.X.Wolf            0.002926
x0.X.Moose          -1.192645
Initial states (x0) defined at t=1

Standard errors have not been calculated. 
Use MARSSparamCIs to compute CIs and bias estimates.

Convergence warnings
 Warning: the  R.(Wolf,Wolf)  parameter value has not converged.
 Warning: the  R.(Moose,Moose)  parameter value has not converged.
 Warning: the  x0.X.Wolf  parameter value has not converged.
 Warning: the  logLik  parameter value has not converged.
 Type MARSSinfo("convergence") for more info on this warning.
\end{Soutput}
\end{Schunk}
